---
title: "Visualization #3"
description: |
  Plotting change in self-regulation scores from 2014 to 2018
author:
  - name: Joanna Wright
    url: https://github.com/jhwrigh23
date: 03-18-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r libraries, echo=TRUE}

library(tidyverse)
library(rio)
library(here)
library(janitor)
library(maps)
library(mapdata)
library(ggtext)
```
<aside>
Liraries used for this post
</aside>

```{r loading-the-data, echo=FALSE, results='hide'}

d <- import(here("data", "ka_2018-19_lookback_report.xlsx")) %>%
  clean_names()

```


```{r cleaning-the-data, echo=FALSE, results='hide'}

d_tidy <- d %>%
  rename(county = oregon_kindergarten_assessment,
         district_id = x2,
         district = x3,
         inst_id = x4,
         inst_name = x5,
         inst_type = x6,
         stu_grp_type = x7,
         stu_grp = x8,
         avg_self_reg = x9) %>%
  select(c(1:9))

d_tidy <- d_tidy %>%
  filter(avg_self_reg != "*")

d_tidy = d_tidy[-1,]

```

The purpose of Visualization 3 is to see how self-regulation scores have changed over the course of the OKA. I chose to compare the recent data (2018-2019) with data from the first wave of the OKA (2014-2015). 

To do so, I first needed to load the 2014-2015 data. I cleaned it in the same way I cleaned my initial (2018-2019) data. Then I joined the two waves. Once the data were joined, I used `pivot_longer` to create  columns for each year so that I could compare across years.

Then I was ready to try plotting: 
 
```{r loading-2014-2015-data, results='hide'}
dfirstwave <- import(here("data", "ka_2013-14_lookback_report.xlsx")) %>%
  clean_names()

dfirstwave1 <- dfirstwave %>%
  rename(county = oregon_kindergarten_assessment,
         district_id = x2,
         district = x3,
         inst_id = x4,
         inst_name = x5,
         inst_type = x6,
         stu_grp_type = x7,
         stu_grp = x8,
         avg_self_reg = x9) %>%
  select(c(1:9))
```

```{r joining}

dfirstwave1 <- dfirstwave1 %>%
  filter(avg_self_reg != "*")

dfirstwave1 = dfirstwave1[-1,]

dfirstwave1$avg_self_reg = as.numeric(dfirstwave1$avg_self_reg)

dfirstwave2 <- dfirstwave1 %>%
  group_by(county) %>%
  summarise(avg_self_reg_county_firstwave = mean(avg_self_reg))

dfirstwave2 <- dfirstwave2 %>%
  mutate(county = str_to_lower(county))

d_tidy <- d_tidy %>%
  mutate(county = str_to_lower(county))

join2 <- left_join(d_tidy, dfirstwave2)

join2 <- join2 %>%
  rename(`2014` = avg_self_reg_county_firstwave,
         `2018` = avg_self_reg)

join2$`2014` = as.numeric(join2$`2014`)
join2$`2018` = as.numeric(join2$`2018`)

join2 <- join2 %>%
  select(1, 9, 10)

```

```{r pivoting}

join2 <- join2 %>%
  pivot_longer(
    cols = 2:3,
    names_to = "Year", 
    values_to = "Score",
  )
```

## Version 1
```{r}
join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  ggplot(aes(county, avg_score)) +
  geom_line(aes(group = county), color = "gray40") +
  geom_point(aes(color = Year)) +
  coord_flip() +
  labs(title = "Self Regulation Score Changes from 2014 to 2018",
       x = NULL,
       y = "Score (on 1-5 scale)")
```

But this is difficult to make sense of, particularly because the arrangement of counties in alphabetical order does nothing to help us understand the data. So, I reordered counties by average score: 

## Version 2
```{r}
join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  filter(county != "all counties") %>%
  ungroup() %>%
  ggplot(aes(fct_reorder(county, avg_score), avg_score)) +
  geom_line(aes(group = county), color = "gray40") +
  geom_point(aes(color = Year), size = 1.5) +
  coord_flip() +
  theme(
    plot.title.position = "plot",
    plot.subtitle = element_markdown()) +
  labs(title = "Self Regulation Score Changes from 2014 to 2018",
       subtitle = "<span style = 'color:#FF6C67;'>2014</span> compared to <span style = 'color:#00C2C6;'>2018 </span>change.",
       x = NULL,
       y = "Score (on 1-5 scale)") + 
  guides(color = "none")

```
But this still isn't super helpful, because counties that saw increases in scores are all mixed up with counties that saw decreases in scores. Instead, I wanted to arrange counties not by average score, but by the change from 2014 to 2018. Hopefully, this would enable the viewer to easily identify counties in which scores may be decreasing, which can inform choices about providing resources to support those locales:

## Version 3
```{r }

join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "avg_score"
  ) %>%
  mutate(change_14_to_18 = c(`2014` - `2018`)) %>%
  pivot_longer(
    cols = 2:3,
    names_to = "Year",
    values_to = "Score"
    ) %>%
  ggplot(aes(fct_reorder(county, change_14_to_18), Score)) +
  geom_point(aes(color = Year), size = 1.5) +
  geom_line(aes(group = county), color = "gray40") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Self Regulation Score Changes from 2014 to 2018",
       x = NULL,
       y = "Score (on 1-5 scale)")
```

That is better. In the next iteration, I tidied it up a bit more (e.g. title case for county names). I also tried playing with the color palette:

## Version 4
```{r }

join2 <- join2 %>%
  mutate(county = str_to_title(county))

join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "avg_score"
  ) %>%
  mutate(change_14_to_18 = c(`2018` - `2014`)) %>%
  pivot_longer(
    cols = 2:3,
    names_to = "Year",
    values_to = "Score"
    ) %>%
  filter(county != "all counties") %>%
  ggplot(aes(fct_reorder(county, -change_14_to_18), Score)) +
  geom_line(aes(group = county), color = "gray40", size = 1) +
  geom_point(aes(color = Year), size = 2.2) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Self Regulation Score Changes from 2014 to 2018",
       subtitle = "Ordered from greatest increase to greatest decrease",
       x = NULL,
       y = "Score (on 1-5 scale)") + 
  scale_color_brewer(palette = "Set1")

```

I also thought the plot would be clearer if the scale colors displayed in the subtitle rather than in a legend:

## Version 5
```{r }

join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "avg_score"
  ) %>%
  mutate(change_14_to_18 = c(`2018` - `2014`)) %>%
  pivot_longer(
    cols = 2:3,
    names_to = "Year",
    values_to = "Score"
    ) %>%
  filter(county != "all counties") %>%
  ggplot(aes(fct_reorder(county, -change_14_to_18), Score, color = Year)) +
  geom_line(aes(group = county), 
            color = "gray40", 
            size = 1) +
  geom_point(size = 2.2) +
  scale_color_manual(
    name = NULL,
    values = c(`2014` = "#377eb8", `2018` = "#e41a1c")) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.subtitle = element_markdown()) +
  labs(title = "Changes in Average Self-Regulation Scores by County",
       subtitle = "<span style = 'color:#377eb8;'>2014</span> to <span style = 'color:#e41a1c;'>2018 </span>",
       x = NULL,
       y = "Score (on 1-5 scale)") + 
  guides(color = "none")

```

Although this works, I wanted to highlight the counties which saw **increases** in scores vs. the counties that saw **decreases** in scores. This is the main point of the plot and should be immediately evident to the viewer. To do this, I added another `geom_line` for only those counties that saw decreases. I also reversed the `fct_reorder` to make it more intuitive, with counties showing increases in scores at the top and counties showing decreases in scores at the bottom:

## Final Version
```{r preview=TRUE, fig.width=3, fig.height=3}

join2 <- join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "avg_score"
  ) %>%
  mutate(change_14_to_18 = c(`2018` - `2014`)) %>%
  pivot_longer(
    cols = 2:3,
    names_to = "Year",
    values_to = "Score"
    ) %>%
  filter(county != "all counties")

join2$change_14_to_18 = as.numeric(join2$change_14_to_18)

join2_decreased <- join2 %>%
  filter(change_14_to_18 < 0)

join2 %>%
  ggplot(aes(fct_reorder(county, change_14_to_18), Score, color = Year)) +
  geom_line(aes(group = county), 
            color = "gray40", 
            size = 1) +
  geom_line(data = join2_decreased, aes(group = county),
            color = "#e41a1c",
            size = 1) + 
  geom_point(size = 2.4) +
  scale_color_manual(
    name = NULL,
    values = c(`2014` = "#6baed6", `2018` = "#084594")) +
  coord_flip() +
  theme_minimal() + 
  theme(
    plot.title.position = "plot",
    plot.subtitle = element_markdown()) +
  labs(title = "Changes in Average Self-Regulation Scores by County",
       subtitle = "<span style = 'color:#6baed6;'>2014</span> to <span style = 'color:#084594;'>2018</span>, with<span style = 'color:#e41a1c;'</span> decreases over time </span>highlighted", 
       x = NULL,
       y = "Score (on 1-5 scale)",
       caption = "Note. These data must be interpreted with caution because\nthey are from only 2 time points (2014 and 2018).\n Some counties may show variation year to year but\nnot necessarily show a strong directional trend across all waves.") + 
  guides(color = "none")

```



