---
title: "Set up"
description: |
  Here is the code for loading and initial cleaning of the data. 
author:
  - name: Joanna Wright
    url: https://github.com/jhwrigh23
date: 03-18-2020
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r libraries}

library(tidyverse)
library(rio)
library(here)
library(janitor)

```

I loaded the Oregon Kindergarten Assessment data by downloading the Excel file from https://www.oregon.gov/ode/educator-resources/assessment/Pages/Kindergarten-Assessment.aspx and then loading it: 

```{r loading the data}

d <- import(here("data", "ka_2018-19_lookback_report.xlsx")) %>%
  clean_names()

```

I did an initial cleaning of the data. This created my base data object, **d_tidy**. I selected my variables of interest and chose to focus specifically on self-regulation. I also filtered out missing data, which brought the number of observations down from 14,443 to 6,160 (many observations did not have self-regulation scores). 

**d_tidy** was further refined in specific ways for each visualization. The code for these iterations of data cleaning is within each Post and can be accessed via my GitHub site in the [_posts](https://github.com/jhwrigh23/class_website/tree/master/_posts) folder. 


```{r cleaning the data, results='hide'}

d_tidy <- d %>%
  rename(county = oregon_kindergarten_assessment,
         district_id = x2,
         district = x3,
         inst_id = x4,
         inst_name = x5,
         inst_type = x6,
         stu_grp_type = x7,
         stu_grp = x8,
         avg_self_reg = x9) %>%
  select(c(1:9))

d_tidy <- d_tidy %>%
  filter(avg_self_reg != "*")

d_tidy = d_tidy[-1,]

str(d_tidy)
d_tidy$avg_self_reg = as.numeric(d_tidy$avg_self_reg)

```

