---
title: "Visualization 1"
description: |
  Mapping self-regulation scores by county
author:
  - name: Joanna Wright
    url: https://github.com/jhwrigh23
date: 02-26-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      fig.height = 7)
```

Libraries for Visualization 1: 
```{r libraries}

library(tidyverse)
library(rio)
library(here)
library(janitor)
library(maps)
library(mapdata)
library(ggtext)
library(usmap)
library(sf)
library(sp)
```

```{r loading the data, results='hide'}

d <- import(here("data", "ka_2018-19_lookback_report.xlsx")) %>%
  clean_names()

```

```{r cleaning the data, results='hide'}

d_tidy <- d %>%
  rename(county = oregon_kindergarten_assessment,
         district_id = x2,
         district = x3,
         inst_id = x4,
         inst_name = x5,
         inst_type = x6,
         stu_grp_type = x7,
         stu_grp = x8,
         avg_self_reg = x9) %>%
  select(c(1:9))

d_tidy <- d_tidy %>%
  filter(avg_self_reg != "*")

d_tidy = d_tidy[-1,]

str(d_tidy)
d_tidy$avg_self_reg = as.numeric(d_tidy$avg_self_reg)

```

The purpose of Plot 1 is to explore differences in average self-regulation scores across counties and represent this geographically. 

To do so, I first needed to load Oregon county map data and then join the OKA dataset with the mapping data.

```{r viz1-joining, results='hide'}

counties <- map_data("county")
or_county <- subset(counties, region == "oregon")

# identifying keys, joining: 

d %>%
  count(oregon_kindergarten_assessment, x2, x3, x4, x5, x6, x7, x8) %>%
  filter(n>1)

or_county %>%
  count(order) %>%
  filter(n>1)

d1 <- d_tidy %>%
  group_by(county) %>%
  summarise(avg_self_reg_county = mean(avg_self_reg))

or_county <- or_county %>%
  rename(county = subregion)

d1 <- d1 %>%
  mutate(county = str_to_lower(county))

join1 <- left_join(d1, or_county)
```

Then I tried creating the map:

```{r viz1-vers1}

join1 %>%
ggplot() + 
    geom_polygon(aes(long, lat, group = group, fill = avg_self_reg_county)) + 
    coord_fixed(1.3) +
    scale_fill_viridis_c(name = "Score (1-5 scale)", 
                         option = "inferno") +
    labs(title = "Average Self Regulation Scores by County in Oregon",
         caption = "Data from https://www.oregon.gov/ode/educator-resources/assessment/Pages/Kindergarten-Assessment.aspx",
         x = NULL,
         y = NULL) +
     theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank())

```

This map would be more informative if the counties were labeled.  

```{r viz1-vers2}

join1 %>%
ggplot() + 
    geom_polygon(aes(long, lat, group = group, fill = avg_self_reg_county)) + 
    coord_fixed(1.3) +
    scale_fill_viridis_c(name = "Score (1-5 scale)", 
                         option = "inferno") +
    labs(title = "Average Self Regulation Scores by County in Oregon",
         x = NULL,
         y = NULL) +
     theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank()) +
  geom_text(data = or_county, aes(long, lat, label = county), size = 2)


```

Beautiful, right?!?! I needed to create centroid data in order to label counties properly. 

```{r viz1-vers3}


getLabelPoint <- function(county) {Polygon(county[c('long', 'lat')])@labpt}
df <- map_data('county', 'oregon')
centroids <- by(df, df$subregion, getLabelPoint)    
centroids2 <- do.call("rbind.data.frame", centroids)  
centroids2$county = rownames(centroids)
names(centroids2) <- c('clong', 'clat', "county")               

join3 <- left_join(join1, centroids2)

join3 %>%
ggplot() + 
  geom_polygon(data = join3, aes(x = long, 
                              y = lat, 
                              group = group), 
               fill = "white", 
               color = "gray") + 
  geom_polygon(aes(long, lat, group = group, fill = avg_self_reg_county)) +
  coord_fixed(1.3) + 
  geom_text(aes(x = clong, y = clat, label = county), color = "darkblue", size = 3) + 
scale_fill_viridis_c(name = "Score (1-5 scale)", 
                         option = "inferno") +
    labs(title = "Average Self Regulation Scores by County in Oregon",
         x = NULL,
         y = NULL) + 
     theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank())

```

While this accomplished what I wanted it to functionally, it doesn't look very nice. The blue labels are difficult to read against this these scale colors. So I tried it a few different ways. 

```{r viz1-vers4}

join3 %>%
ggplot() + 
  geom_polygon(data = join3, aes(x = long, 
                              y = lat, 
                              group = group), 
               fill = "white", 
               color = "gray") + 
  geom_polygon(aes(long, lat, group = group, fill = avg_self_reg_county)) +
  coord_fixed(1.3) + 
  geom_text(aes(x = clong, y = clat, label = county), color = "white", size = 1.75) + 
scale_colour_brewer() +
    labs(title = "Average Self Regulation Scores by County in Oregon",
         x = NULL,
         y = NULL) + 
     theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank())


```

Some final refining: 
```{r viz1-vers5}

join3 <- join3 %>%
  mutate(county = str_to_title(county))

join3 %>%
ggplot() + 
  geom_polygon(data = join3, aes(x = long, 
                              y = lat, 
                              group = group), 
               fill = "white", 
               color = "gray") + 
  geom_polygon(aes(long, lat, group = group, fill = avg_self_reg_county)) +
  coord_fixed(1.3) + 
  geom_text(aes(x = clong, y = clat, label = county), 
            color = "white", 
            size = 1.75, 
            check_overlap = TRUE) + 
scale_colour_brewer() +
  scale_fill_continuous(name = "Average score (1-5 scale)") +
    labs(title = "Average Self Regulation Scores by County in Oregon",
         x = NULL,
         y = NULL) + 
     theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank(),
        legend.position = "bottom")

```
