---
title: "Visualization 2"
description: |
  Comparing self-regulation scores by race
author:
  - name: Joanna Wright
    url: https://github.com/jhwrigh23
date: 03-09-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```

```{r}

library(tidyverse)
library(rio)
library(here)
library(janitor)
library(gt)
library(ggrepel)

```


```{r loading-the-data}

d <- import(here("data", "ka_2018-19_lookback_report.xlsx")) %>%
  clean_names()

```

```{r cleaning-the-data}

d_tidy <- d %>%
  rename(county = oregon_kindergarten_assessment,
         district_id = x2,
         district = x3,
         inst_id = x4,
         inst_name = x5,
         inst_type = x6,
         stu_grp_type = x7,
         stu_grp = x8,
         avg_self_reg = x9) %>%
  select(c(1:9))

d_tidy <- d_tidy %>%
  filter(avg_self_reg != "*")

d_tidy = d_tidy[-1,]

str(d_tidy)
d_tidy$avg_self_reg = as.numeric(d_tidy$avg_self_reg)

```

The purpose of Plot 2 is to explore variation in self-regulation scores by race in Oregon. 

Because the student group variable included both racial and gender categories, the first step was to filter for only race categories. 

```{r viz2}

d2 <- d_tidy %>%
  filter(stu_grp == "Asian" |
           stu_grp == "White" |
           stu_grp == "Multi-Racial" |
           stu_grp == "Native Hawaiian/Pacific Islander" |
           stu_grp == "Hispanic/Latino" |
           stu_grp == "Black/African American" |
           stu_grp == "American Indian/Alaska Native")

```

First, just to get an overall sense of the data, I plotted average scores by race. 

```{r}

d2 %>%
  group_by(stu_grp) %>%
  summarise(avg_self_reg_race = mean(avg_self_reg)) %>%
  ggplot(aes(fct_reorder(stu_grp, avg_self_reg_race), avg_self_reg_race)) +
  geom_point() +
  coord_flip() + 
  labs(title = "Average self-regulation by race",
       x = NULL,
       y = "Self-regulation score")

```

To explore the data further, I  plotted distributions in a couple of difference ways.  

```{r viz2-alternative}

d2 %>%
ggplot(aes(avg_self_reg, fct_reorder(stu_grp, avg_self_reg))) +
  ggridges::geom_density_ridges(color = "white", 
                                fill = "cornflowerblue", 
                                alpha = 0.8) + 
  theme_minimal(base_size = 10) + 
  scale_x_continuous("Self Regulation Scores") +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(color = "gray80")) +
  labs(title = "Self-Regulation Scores by Race",
       x = NULL,
       y = NULL)

d2 %>%
  ggplot(aes(avg_self_reg, fct_reorder(stu_grp, avg_self_reg))) +
  geom_violin(fill = "cornflowerblue")

d2 %>%
  ggplot(aes(avg_self_reg, fct_reorder(stu_grp, avg_self_reg), fill = stu_grp)) +
  geom_violin() +
  labs(title = "Distribution of self-regulation scores by race",
       x = "Score (1-5 scale)",
       y = NULL) +
  theme_minimal() + 
  scale_color_brewer(palette = "Accent") +
  theme(plot.title = element_text(hjust = 1)) +
  theme(legend.position = "none") 
  
```

To put these numbers in context, it is important to know that some student groups are represented much more strongly than others. 

```{r}

d2 %>%
  group_by(stu_grp) %>%
  count()
             
```

I decided to look at difference in self-regulation scores between the largest student groups represented (White, Hispanic/Latino).  

```{r}

d2 %>%
  filter(stu_grp == "White" |
           stu_grp == "Hispanic/Latino") %>%
  group_by(stu_grp, county) %>%
  summarise(avg_self_reg_race = mean(avg_self_reg)) %>%
  pivot_wider(
    names_from = stu_grp,
    values_from = avg_self_reg_race
  ) %>%
  mutate(white_hisp_diff = c(White - `Hispanic/Latino`)) %>%
  ggplot(aes(county, white_hisp_diff)) +
  geom_point() +
  coord_flip()

```

Refining the plot to highlight counties in which there is the largest score difference between White and Hispanic/Latino students.  

```{r}

d2 %>%
  filter(stu_grp == "White" |
           stu_grp == "Hispanic/Latino") %>%
  group_by(stu_grp, county) %>%
  summarise(avg_self_reg_race = mean(avg_self_reg)) %>%
  pivot_wider(
    names_from = stu_grp,
    values_from = avg_self_reg_race
  ) %>%
  mutate(white_hisp_diff = c(White - `Hispanic/Latino`)) %>%
  filter(white_hisp_diff != "NA") %>%
  ggplot(aes(fct_reorder(county, white_hisp_diff), white_hisp_diff)) +
  geom_point(col = "darkred") +
  coord_flip() + 
  gghighlight::gghighlight(white_hisp_diff < -.25 |
                             white_hisp_diff > .25,
                           unhighlighted_colour = alpha("steelblue", 0.4)) + 
  geom_point(col = "darkred", size = 2.5) + 
  labs(title = "Differences Self-Regulation Scores by Race in Oregon",
       subtitle = "Disparities in scores between White and Hispanic/Latino students",
       x = NULL,
       y = "Difference in scores")
  
```

How can this be improved? 

```{r}

d2 %>%
  filter(stu_grp == "White" |
           stu_grp == "Hispanic/Latino") %>%
  group_by(stu_grp, county) %>%
  summarise(avg_self_reg_race = mean(avg_self_reg)) %>%
  pivot_wider(
    names_from = stu_grp,
    values_from = avg_self_reg_race
  ) %>%
  mutate(white_hisp_diff = c(White - `Hispanic/Latino`)) %>%
  filter(white_hisp_diff != "NA") %>%
  ggplot(aes(fct_reorder(county, white_hisp_diff), white_hisp_diff, 
         fill = white_hisp_diff > 0)) +
  geom_bar(stat = 'identity') +
  coord_flip() + 
  labs(title = "Differences Self-Regulation Scores by Race in Oregon",
       subtitle = "Disparities in scores between White and Hispanic/Latino students",
       x = NULL,
       y = "Difference in scores")

```

That's easier to read, but it's still hard to tell what it means. I needed to find a way to communicate that the blue represents White student scores being higher than Hispanic, and the red represents Hispanic student scores being higher than White. 

```{r}

d2 %>%
  filter(stu_grp == "White" |
           stu_grp == "Hispanic/Latino") %>%
  group_by(stu_grp, county) %>%
  summarise(avg_self_reg_race = mean(avg_self_reg)) %>%
  pivot_wider(
    names_from = stu_grp,
    values_from = avg_self_reg_race
  ) %>%
  mutate(white_hisp_diff = c(White - `Hispanic/Latino`)) %>%
  filter(white_hisp_diff != "NA") %>%
  ggplot(aes(fct_reorder(county, white_hisp_diff), white_hisp_diff, 
         fill = white_hisp_diff > 0)) +
  geom_bar(stat = 'identity') +
  scale_fill_viridis_d(name = "Average White student scores higher than Hispanic/Latino") +
  coord_flip() + 
  labs(title = "Differences Self-Regulation Scores by Race in Oregon",
       subtitle = "Disparities in scores between White and Hispanic/Latino students",
       x = NULL,
       y = "Difference in scores") +
  theme_dark() + 
  theme(legend.position = "bottom")


```


```{r fig.height=3, fig.width=4}

d2 %>%
  filter(stu_grp == "White" |
           stu_grp == "Hispanic/Latino") %>%
  group_by(stu_grp, county) %>%
  summarise(avg_self_reg_race = mean(avg_self_reg)) %>%
  pivot_wider(
    names_from = stu_grp,
    values_from = avg_self_reg_race
  ) %>%
  mutate(white_hisp_diff = c(White - `Hispanic/Latino`)) %>%
  filter(white_hisp_diff != "NA") %>%
  ggplot(aes(fct_reorder(county, white_hisp_diff), white_hisp_diff, 
         fill = white_hisp_diff > 0)) +
  geom_bar(stat = 'identity') +
  scale_fill_viridis_d(name = "Average White student scores higher than Hispanic/Latino") +
  coord_flip() + 
  labs(title = "Differences Self-Regulation Scores by Race in Oregon",
       subtitle = "Disparities in scores between White and Hispanic/Latino students",
       x = NULL,
       y = "Difference in scores") +
  theme_dark() + 
  theme(legend.position = "bottom")

```

This is getting pretty close! I'd still like to change the caption to reduce cognitive load - make it  super clear what data are depicted. 

