---
title: "Set up"
description: |
  Here is the code for loading and initial cleaning of the data. 
author:
  - name: Joanna Wright
    url: https://github.com/jhwrigh23
date: 03-10-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r libraries}

library(tidyverse)
library(rio)
library(here)
library(janitor)

```

I loaded the Oregon Kindergarten Assessment data by downloading the Excel file from https://www.oregon.gov/ode/educator-resources/assessment/Pages/Kindergarten-Assessment.aspx and then loading it: 

```{r loading the data}

d <- import(here("data", "ka_2018-19_lookback_report.xlsx")) %>%
  clean_names()

```

I did an initial cleaning of the data, to create a data object I could use and further refine for each visualization. I selected my variables of interest and chose to focus specifically on self-regulation. I also filtered out missing data, which brought the number of observations down from 14,443 to 6,160 (many observations did not have self-regulation scores). 

```{r cleaning the data}

d_tidy <- d %>%
  rename(county = oregon_kindergarten_assessment,
         district_id = x2,
         district = x3,
         inst_id = x4,
         inst_name = x5,
         inst_type = x6,
         stu_grp_type = x7,
         stu_grp = x8,
         avg_self_reg = x9) %>%
  select(c(1:9))

d_tidy <- d_tidy %>%
  filter(avg_self_reg != "*")

d_tidy = d_tidy[-1,]

str(d_tidy)
d_tidy$avg_self_reg = as.numeric(d_tidy$avg_self_reg)

```

